# Introduction

```{r setup, include=FALSE}
WORDS_TO_IGNORE <- c("pak", "EWAS", "methylation")
source("knitr-options.R")
source("spelling-check.R")
```

## Main motivation for developing {bigstatsr} and {bigsnpr}

The main motivation was for me to be able to run all my analyses within `r icon::fa_r_project()`.
I was frustrated by having to use all these different software, with different input formats, and requiring text files for parameters.
This made it hard for me to build a chain of analyses, to perform some exploratory analyses, or to use familiar  `r icon::fa_r_project()` packages.
Also, I wanted to develop new methods, which seemed very hard to do without using a simple matrix-like format.

Thus I started developing  `r icon::fa_r_project()` package {bigsnpr} at the beginning of my thesis.
At some point, I realized that many functions (to perform e.g. [GWAS](https://en.wikipedia.org/wiki/Genome-wide_association_study), [PCA](https://en.wikipedia.org/wiki/Principal_component_analysis), summary statistics) were not really specific to genotype data.
Indeed, a TWAS or an EWAS are not conceptually very different from a GWAS; one can also perform PCA on e.g. DNA methylation data.
Therefore I decided to move all these functions that could be used on any data stored as a matrix and build a new `r icon::fa_r_project()` package, {bigstatsr}.
This is why there are two packages, where {bigstatsr} can basically be used by any field using matrices, while {bigsnpr} provides some tools rather specific to genotype data, largely building on top of {bigstatsr}.

:::: {.infobox .info}

To know which function belongs to which package: functions starting with `big_` belongs to {bigstatsr} while functions starting with `snp_` or `bed_` belongs to {bigsnpr}.

::::


## Installation

Both packages are available on CRAN, so you can use `install.packages()`:

```{r, eval=FALSE}
install.packages("bigstatsr")
install.packages("bigsnpr")
```

To install the latest versions (from GitHub), you can use {remotes}:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("privefl/bigstatsr")
remotes::install_github("privefl/bigsnpr")
```

As an alternative to {remotes}, you can also try [{pak}](https://github.com/r-lib/pak), which gives more information and solve other issues.


## Correct spellings

<blockquote class="twitter-tweet" data-lang="en" align="center" width="45%"><p lang="en" dir="ltr">A friendly reminder:<br>The correct spelling is<br> - bigstatsr -- not bigstatr / BIGstatsR<br> - bigsnpr -- not BIGsnpR / bigSNPr<br> - pcadapt -- not PCAdapt<br> - LDpred -- not LDPred<br><br>Thank you<br>The kittens thank you too <a href="https://t.co/S8wyE4G6BG">pic.twitter.com/S8wyE4G6BG</a></p>&mdash; Florian Privé (remote postdoc) (@privefl) <a href="https://twitter.com/privefl/status/1329109034218237953?ref_src=twsrc%5Etfw">November 18, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 
