# Population structure

```{r setup, include=FALSE}
WORDS_TO_IGNORE <- c("pca", "loadings")
source("knitr-options.R")
source("spelling-check.R")
```

## Principal Component Analysis (PCA)

PCA on the genotype matrix can be used to capture population structure.
PCA can capture different kinds of structure [@prive2020efficient]:

- population structure (what we want),
- LD structure, when there are two many correlated variants and not enough population structure (see [this vignette](https://privefl.github.io/bigsnpr/articles/how-to-PCA.html)),
- relatedness structure, when there are related individuals that can cluster together in later PCs,
- noise, basically just circles when looking at PC scores.

Population structure is the second main topic of my research work (after polygenic scores).

- In @prive2017efficient, I introduced an algorithm to compute PCA for a `bigSNP` object while accounting for the LD problem by using clumping (not pruning, cf. [this vignette](https://privefl.github.io/bigsnpr/articles/pruning-vs-clumping.html)) and an *automatic* removal of long-range LD regions.

- In @prive2020performing, I improved an algorithm to detect variants associated with population structure.

- In @prive2020efficient, I extended the package to be also able to run PCA directly on PLINK bed files with a small percentage of missing values, and investigated best practices for PCA in more detail.

- In @prive2021high and @prive2021using, I showed how to use PCA for ancestry inference, including grouping individuals in homogeneous ancestry groups, and inferring ancestry proportions from allele frequencies only (see [this vignette](https://privefl.github.io/bigsnpr/articles/ancestry.html)).


## Example {#exo-pca}

There are many steps to properly perform a PCA, which I will try to detail in the following exercise.
**You can find another example in [this vignette](https://privefl.github.io/bigsnpr/articles/bedpca.html).**

Let us reuse the data prepared in \@ref(exo-preprocessing).

First, let us get an idea of the relatedness in the data using
```{r, results='hide'}
library(bigsnpr)
(NCORES <- nb_cores())
plink2 <- download_plink2("tmp-data")
rel <- snp_plinkKINGQC(plink2, "tmp-data/GWAS_data_sorted_QC.bed",
                       thr.king = 2^-4.5, make.bed = FALSE, ncores = NCORES)
```

```{r}
hist(log2(rel$KINSHIP), "FD")
```

:::: {.infobox .info}
When computing relatedness with KING, it may be needed to filter out some variants that are highly associated with population structure, e.g. as performed in the UK Biobank [@bycroft2018uk]. For example, see [this code](https://github.com/privefl/bigsnpr-extdoc/blob/main/example-code/2-rel-and-pca.R).
::::

Relatedness should not be a major issue here. Let us now compute PCs.

```{r}
obj.bed <- bed("tmp-data/GWAS_data_sorted_QC.bed")
obj.svd <- runonce::save_run(
  bed_autoSVD(obj.bed, k = 12, ncores = NCORES),
  file = "tmp-data/PCA_GWAS_data.rds")
```

```{r, fig.asp=0.8}
plot(obj.svd)
plot(obj.svd, type = "scores", scores = 1:12, coeff = 0.5)
```

There is some population structure (maybe up to 6 PCs). You should also check loadings to make sure there is no LD structure (peaks on loadings):

```{r}
plot(obj.svd, type = "loadings", loadings = 1:8, coeff = 0.5)
```

No peaks, but loadings of PC4 and PC5 are a bit odd.

If you expect the individuals to mostly come from one population, you can simply use a robust distance to identify a homogeneous subset of individuals, then look at the histogram of log-distances to choose a threshold based on visual inspection (here I would probably choose 4.5).

```{r}
PC <- predict(obj.svd)
ldist <- log(bigutilsr::dist_ogk(PC))
hist(ldist, "FD")
```
```{r, message=FALSE}
library(ggplot2)
devtools::source_url(
  "https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")

plot_grid2(plotlist = lapply(1:4, function(k) {
  k1 <- 2 * k - 1
  k2 <- 2 * k
  qplot(PC[, k1], PC[, k2], color = ldist, size = I(2)) +
    scale_color_viridis_c() +
    theme_bigstatsr(0.6) +
    labs(x = paste0("PC", k1), y = paste0("PC", k2), color = "log-distance") +
    coord_equal()
}), nrow = 2, legend_ratio = 0.2, title_ratio = 0)
```

It would be nice if we could get a better sense of the ancestry of these individuals.
To achieve this, we can project this data onto the PCA space of many known population groups defined in @prive2021using.

```{r}
all_freq <- bigreadr::fread2(
  runonce::download_file(
    "https://figshare.com/ndownloader/files/38019027",  # for the tutorial (46 MB)
    # "https://figshare.com/ndownloader/files/31620968",  # for real analyses (849 MB)
    dir = "tmp-data", fname = "ref_freqs.csv.gz"))

projection <- bigreadr::fread2(
  runonce::download_file(
    "https://figshare.com/ndownloader/files/38019024",  # for the tutorial (44 MB)
    # "https://figshare.com/ndownloader/files/31620953",  # for real analyses (847 MB)
    dir = "tmp-data", fname = "projection.csv.gz"))

correction <- c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099,
                1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
```


```{r, warning=FALSE, message=FALSE}
library(dplyr)
matched <- obj.bed$map %>%
  transmute(chr = chromosome, pos = physical.pos, a1 = allele1, a0 = allele2) %>% 
  mutate(beta = 1) %>%
  snp_match(all_freq[1:5]) %>%
  print()

# further subsetting on missing values
counts <- bed_counts(obj.bed, ind.col = matched$`_NUM_ID_.ss`, ncores = NCORES)
# hist(nb_na <- counts[4, ])
ind <- which(counts[4, ] < (nrow(obj.bed) * 0.05))        

# project individuals (divided by 2) onto the PC space
PROJ <- as.matrix(projection[matched$`_NUM_ID_`[ind], -(1:5)])
all_proj <- apply(sweep(PROJ, 2, correction / 2, '*'), 2, function(x) {
  bed_prodVec(obj.bed, x, ind.col = matched$`_NUM_ID_.ss`[ind], ncores = NCORES,
              # scaling to get G if beta = 1 and (2 - G) if beta = -1
              center = 1 - matched$beta[ind],
              scale = matched$beta[ind])
})
```

We can then assign individuals to the closest center:
```{r}
all_centers <- crossprod(as.matrix(all_freq[matched$`_NUM_ID_`[ind], -(1:5)]), PROJ)
all_sq_dist <- apply(all_centers, 1, function(one_center) {
  rowSums(sweep(all_proj, 2, one_center, '-')^2)
})

THR <- 0.005  # you can adjust this threshold
thr_sq_dist <- max(dist(all_centers)^2) * THR / 0.16

group <- colnames(all_freq)[-(1:5)]
group[group %in% c("Scandinavia", "United Kingdom", "Ireland")]   <- "Europe (North West)"
group[group %in% c("Europe (South East)", "Europe (North East)")] <- "Europe (East)"

cluster <- group[
  apply(all_sq_dist, 1, function(x) {
    ind <- which.min(x)
    if (x[ind] < thr_sq_dist) ind else NA
  })
]

table(cluster, exclude = NULL)  # no NA -> all assigned
```

```{r}
plot_grid2(plotlist = lapply(1:4, function(k) {
  k1 <- 2 * k - 1
  k2 <- 2 * k
  qplot(PC[, k1], PC[, k2], color = cluster, size = I(2)) +
    theme_bigstatsr(0.6) +
    labs(x = paste0("PC", k1), y = paste0("PC", k2), color = "Assigned\npopulation") +
    coord_equal()
}), nrow = 2, legend_ratio = 0.25, title_ratio = 0)
```

These are mostly European individuals.
PC4-PC6 are definitively a bit odd. 
