# Population structure

```{r setup, include=FALSE}
WORDS_TO_IGNORE <- c("pca")
source("knitr-options.R")
source("spelling-check.R")
library(bigsnpr)
```

## Principal Component Analysis (PCA)

PCA on the genotype matrix can be used to capture population structure.
PCA can capture different kinds of structure [@prive2020efficient]:

- population structure (what we want)

- LD structure, when there are two many correlated variants and not enough population structure (see [this vignette](https://privefl.github.io/bigsnpr/articles/how-to-PCA.html))

- relatedness structure, when there are related individuals that can cluster together in later PCs

- noise, basically just circles when looking at PC scores

In @prive2017efficient, I have introduced an algorithm to compute PCA for a `bigSNP` object while accounting for the LD problem via an automatic removing of long-range LD regions.
In @prive2020efficient, I have extended the package to be also able to run PCA directly on PLINK bed files with a few missing values, and investigated best practices for PCA in more detail.
In @prive2021high, I have shown how to use PCA for ancestry inference.

There are many steps to compute properly a PCA analysis, which I will try to detail in the following exercise.


## Exercise: population structure {#exo-pca}

**You can find another example in [this vignette](https://privefl.github.io/bigsnpr/articles/bedpca.html).**

Let us reuse the data prepared in \@ref(exo-preprocessing).

First, let's get an idea of the relatedness in the data using
```{r, results='hide'}
library(bigsnpr)
plink2 <- download_plink2("tmp-data")
rel <- snp_plinkKINGQC(plink2, "tmp-data/GWAS_data_sorted_QC.bed",
                       thr.king = 2^-4.5, make.bed = FALSE, ncores = nb_cores())
```

```{r}
hist(log2(rel$KINSHIP), "FD")
```

:::: {.infobox .info}
When computing relatedness with KING, it may be needed to filter out some variants that are highly associated with population structure, e.g. as performed in the UK Biobank [@bycroft2018uk]. [TODO: also export my code for iPSYCH]
::::

```{r}
obj.bed <- bed("tmp-data/GWAS_data_sorted_QC.bed")
obj.svd <- runonce::save_run(
  bed_autoSVD(obj.bed, k = 20, ncores = nb_cores()),
  file = "tmp-data/PCA_GWAS_data.rds")
```


```{r, fig.asp=0.8}
plot(obj.svd)
plot(obj.svd, type = "scores", scores = 1:8, coeff = 0.6)
```

There is a bit of population structure (it seems at least 6 PCs). It would be better if could get a better sense of the ancestry of these individuals; let us project this data to the 1000G data and infer ancestry as described in @prive2021high.

```{r}
bed.ref <- bed(download_1000G("tmp-data"))
proj <- runonce::save_run(
  bed_projectPCA(bed.ref, obj.bed, k = 25, ncores = nb_cores()),
  file = "tmp-data/proj-to-1000G.rds")
```

```{r}
PC.ref <- predict(proj$obj.svd.ref)
proj2 <- proj$OADP_proj
fam2 <- bigreadr::fread2(sub_bed(bed.ref$bedfile, ".fam2"))
```

```{r, warning=FALSE, out.width = "80%"}
library(ggplot2)
source("https://raw.githubusercontent.com/privefl/paper4-bedpca/master/code/plot_grid2.R")
plot_grid2(plotlist = lapply(1:9, function(k) {
  k1 <- 2 * k - 1
  k2 <- 2 * k
  qplot(PC.ref[, k1], PC.ref[, k2], color = fam2$`Super Population`, size = I(2)) +
    geom_point(aes(proj2[, k1], proj2[, k2]), color = "black", alpha = 0.1) +
    theme_bigstatsr(0.5) +
    labs(x = paste0("PC", k1), y = paste0("PC", k2), color = "Ref Pop") +
    coord_equal()
}), nrow = 3)
```

These are clearly mostly European individuals.

```{r}
ldist <- log(bigutilsr::dist_ogk(proj2))
lims <- bigutilsr::hist_out(ldist)$lim
hist(ldist, "FD"); abline(v = lims, col = "red") 
sum(ldist > lims[2])
```

There are `r sum(ldist > lims[2])` outlier individuals that you could remove.


